Efficient Computation of Extended Sources

### 2021-06-17 repro/papers/sseprecond/paper.tex
includes cg vs pcg plots, but no text

Next observe that in view of the relation \ref{eqn:tracejump20}, and
the downgoing nature of the ray system carrying the high frequency
energy in $(p^+,\bv^+)$, the field $(\tilde{p}^-,\tilde{\bv}^-)$ is
actually the restriction to $z<z_r$ of the anti-causal solution of \ref{eqn:awepm}
with $z_s$ replaced by $z_r$, zero constitutive defect, and vertical
load given by the jump in pressure at $z=z_r$ - for this field, use
the same notation. Continuity of vertical
velocity $\tilde{v}^-_z$ at $z=z_r$ implies that the vertical load is
\[
  f_r = -[\tilde{p}^-]|_{z=z_r} =-(\lim_{z\rightarrow
    z_r^+}\tilde{p}^- - \lim_{z\rightarrow
    z_r^-}\tilde{p}^-)
\]
\[
  \approx 2 P_r \tilde{p}^- = 2 P_r p^+
\]
(from the definition \ref{eqn:defsamp}, $P_r$ is the limit from the
left). Thus
\[
  P_s \tilde{v^-_z} \approx V^-_{z_r,z_s}(2 P_rp^+) \approx
  2V^-_{z_r,z_s}S^+_{z_r,z_s}h_s.
\]
so
\[
  h_s \approx -2 P_s v^+_z \approx -2 P_s \tilde{v}^-_z \approx
  -4V^-_{z_r,z_s}S^+_{z_r,z_s}h_s
\]
Combine this observation with \ref{eqn:tildevtohsubs} to obtain
\[
 -4  V^-_{z_r,z_s} S^+_{z_s,z_r}  \approx  I,
\]
This relation combines with the identity \ref{eqn:trtrcomp} to
yield the first main result of this section:
\begin{eqnarray}
  \label{eqn:approxinv}
  (V^+_{z_s,z_r})^T S^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
  (S^+_{z_s,z_r})^T V^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
  V^+_{z_s,z_r} (S^+_{z_s,z_r})^T & \approx & \frac{1}{4}I, \nonumber\\
  S^+_{z_s,z_r} (V^+_{z_s,z_r})^T & \approx & \frac{1}{4}I.\\
\end{eqnarray}.
The second equation is simply the transpose of the first, and the
last two follow by by an exactly analogous argument using time
reversal and interchange of the roles of $z_s$ and$z_r$.

@@

Figure \ref{fig:dinvhshh0} shows the approximate inversion (via the
first equation in display \ref{eqn:approxinv}) of the
pressure gather shown in Figure \ref{fig:drecphh0}, inverted in the homogeneous model
used to generate the data. The result is a scaled, dip-filtered
version of the pressure source shown in Figure \ref{fig:dhshh0}.
The dip filter effect is unavoidable and is caused by the aperture
limitation of the acquisition geometry: the steeper dips in the source
gather \ref{fig:dhshh0} do not contribute to the data, and are not
recovered in the inversion.

@@

The second feature is a decoupling of the models used (implicitly or
otherwise) in simulation and those used in inversion: it is possible
to {\em accurately invert pressure data for pressure source using the
  wrong model}, in the sense that the inverted source will
generate an accurate recovery of the input data re-simulated in the
same (wrong) model. For example, inversion of the data of the previous
example (Figure \ref{fig:drecplh0}) {\em in the homogeneous model}
results in an inverted source gather at $z=z_s$ (Figure \ref{fig:dinvhslh0}
that is very far from a dip-filtered version of the source used
generate the data \ref{fig:dhshh0}. However, this result is an
accurate inversion: re-simulation {\em using the same (homogeneous)
  model as used in the inversion} results in accurate recovery of the
input pressure gather (Figure \ref{fig:drecplh0}). The difference is
shown on the same color scale in Figure \ref{fig:ddiffrecplh0}.

@@

Relation \ref{eqn:tracejump10} characterizes
$\Lambda^+_{z_s}$ as connecting the pressure and velocity components
of downgoing fields restricted to $z=z_s$. That is, the pressure
source gather
$h_s=\Lambda^{+}_{z_s}P_s p^+ = -2
P_{z_s}v_z$, displayed in Figure \ref{fig:dhshh0}, is the image of the
pressure gather in Figure \ref{fig:dsrcphh0} under
$\Lambda^{+}_{z_s}$.
The pressure gather
\ref{fig:dfwdphh0} is the image of this pressure source gather under
$S^+_{z_s,z_r}$. The corresponding vertical velocity gather
(Figure \ref{fig:dfwdvzhh0}) is $-1/2$ times $\Lambda^+_{z_r}P_r
p$. Therefore scaling the data in Figure \ref{fig:dfwdvzhh0} by
$-2$ produces
$Lambda^+_{z_r}S^+_{z_s,z_r}\Lambda^{+}_{z_s}P_s p^+$. On the other
hand, figure \ref{fig:daltvzhh0} shows the result of applying
$V^+_{z_s,z_r}$ to $f_s=-2P_s p^+$. Therefore scaling the gather in
Figure \ref{fig:daltvzhh0} by $-\frac{1}{2}$ produces $V^+_{z_s,z_r}P_zp^+$.
Since the data in Figures \ref{fig:dfwdvzhh0} and \ref{fig:daltvzhh0}
are essentially identical, the relation \ref{eqn:sv} holds for this
example.

As shown in the last section, $4(V_{z_s,z_r}^+)^T$ is approximately
inverse to $S^{+}_{z_s,z_r}$. Therefore, transposing both sides of
equation \ref{eqn:sv} and using \ref{eqn:approxinv}, obtain
\begin{equation}
  \label{eqn:almostunitary}
  4(V_{z_s,z_r}^+)^TS^+_{z_s,z_r} = [ (\Lambda^+_{z_r})^T
  (S^{+}_{z_s,z_r})^T(\Lambda^+_{z_s})^T]S^{+}_{z_s,z_r} \approx I.
\end{equation}

The remarkable feature of the identity \ref{eqn:almostunitary} is that
it exhibits an approximate right inverse of $S^+$ as an adjoint with
respect to a weighted inner product - or it would, if the operators
$(\Lambda^+)$ were symmetric positive definite. As noted earlier,
these operators are only approximately symmetric, though they are
positive semi-definite. That is not a great obstacle, however:
symmetrizing them in the obvious way commits a negligible error, of
the sort that this paper already neglects wholesale. That is,
\begin{equation}
  \label{eqn:unitary}
  [ \frac{1}{2}((\Lambda^+_{z_r})^T+ \Lambda^+_{z_r})
  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_s})^T+
  \Lambda^+_{z_s})]S^{+}_{z_s,z_r} \approx I.
\end{equation}
The symmetrized $\Lambda$ operators are at least positive
semi-definite, hence define (at least) semi-norms.

Similar relations have been derived for other scattering operators,
and have been used to accelerate iterative solutions of inverse
scatering problems: \cite{DafniSymes:SEG18b} review some of this
literature.

@@

For convenience, in this section write $S$ in place of
$S^+_{z_s,z_r}$. Also abbreviate the symmetrized $\Lambda$ operators
using notation suggesting weight
operators in model and data spaces:
\begin{eqnarray}
  W_m^{-1}&=& \frac{1}{2}((\Lambda^+_{z_s})^T+
              \Lambda^+_{z_s}),\nonumber \\
  W_d &=& \frac{1}{2}((\Lambda^+_{z_r})^T+ \Lambda^+_{z_r}).
          \label{eqn:wdef}
\end{eqnarray}
The identification of the symmetrized $\Lambda^+_{z_s}$ as the inverse
of another operator $W_m$ is formal, since the former operator is
likely to have null (or nearly-null) vectors due to aperture-related
amplitude loss. I will proceed as if this were sensible, since the
mathematical issues can be defused and the only the application of
$W_m^{-1}$ is required in the algorithms described here.

@@

Note that the normal operator appearing on the left-hand side of
\ref{eqn:norm0} is not an approximate identity, due to the presence of
the regularization term: the spectrum increases in spread with
increasing $\alpha$, leading to slower convergence. Fortunately for
the present setting, the operators $W_m^{-1}$, $A$, and $W_m$
approximately commute (they are scalar {\em pseudodifferential}, once
the difficulties with the definition of $W_m$, mentioned above, are
taken care of), so that $A^{\dagger} \approx A^T$. Therefore
\begin{equation}
  \label{eqn:normapprox}
  S^{\dagger}S + \alpha^2 A^{\dagger}A \approx I + \alpha^2A^TA
\end{equation}
Recall that $A$ is simply multiplication by the Euclidean distance to
the physical source point $\bx_s$: $A u (\bx) = |\bx-\bx_s|u(\bx),
A^TAu(\bx) = |\bx-\bx_s|^2u(\bx)$. So the equation $(I+\alpha^2
A^TA)u=b$ is trivial to solve, and this is a key characteristic of a
good preconditioner. However this observation must be combined with
the wegithed norm structure.

Rewrite the normal equation \ref{eqn:norm} as
\begin{equation}
  \label{eqn:norm1}
  W_m^{-1}(S^TW_dS + \alpha^2 A^TW_mA)h = W_m^{-1}S^TW_md 
\end{equation}
Since $W_m$ is self-adjoint and positive definite, the common factor on both sides of \ref{eqn:norm1} can be re-written as
\begin{equation}
  \label{eqn:normpart}
  Nh = (S^*S + \alpha^2 A^*A)h = S^*d 
\end{equation}
in which $S^*, A^*$ are the adjoints with the original (Euclidean)
inner product in the domains but the weighted inner product in data
space:
\begin{eqnarray}
  \label{eqn:sadj}
  S^* &=& S^T W_d,\\
  A^* &=& A^T W_m.
\end{eqnarray}
Note the $S^*S$ and $A^*A$ are symmetric in the Euclidean sense, so
equation \ref{eqn:normpart} is a symmetric positive (semi-)definite
linear system, just the sort of thing for which the 
The Preconditioned Conjugate Gradient (``PCG'') algorithm was
designed. PCG for solution
of equation \ref{eqn:normpart} with preconditioner $M$is usually
written as Algorithm 1 (see for example \cite{Golub:2012}):

@@

Computations of $\Lambda^{\pm}_{z_s}$ and its transpose are clearly critical steps in an
implementation of the PCG algorithm outlined in the preceding section.
Direct computation of the pressure-to-source operator $\Lambda^{\pm}_{z_s}$, for instance by solving
\ref{eqn:awepm} and reading off $P_sv^{\pm}_{z}$, turns out to be
numerically ill-behaved. The relation \ref{eqn:snull} provides and
alternative approach, taking advantage of the accurate approximate
inverse to $S^+_{z_s,z_r}$ constructed above. The first row of
\ref{eqn:snull} is given explicitly in \ref{eqn:snull1}, which I
repeat for the reader's convenience:
\begin{equation}
  \label{eqn:lamidea}
  \Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T f_s \approx  -\frac{1}{2} S^+_{z_s,z_r}\Lambda^+_{z_s}f_s.
\end{equation}

The approximate inverse construction for $S^++_{z_s,z_r}$ permits (approximate)
solution of this equation for $\Lambda^+_{z_s}f_s$: apply
$4(V^{+}_{z_s,z_r})^T$ to both sides of equation \ref{eqn:lamidea} and
use the first equation in the list \ref{eqn:approxinv} to get
\begin{equation}
  \label{eqn:lamident}
  \Lambda^+_{z_s} \approx -8(V^{+}_{z_s,z_r})^T\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T.
\end{equation}
This identity is the second major result of this section: it shows how
to compute that action of $\Lambda^+_{z_s}$ by propagating the input
pressure trace, identified as a source for the velocity evolution,
forward in time from $z_s$ to $z_r$
reading off the pressure trace on $z=z_r$, identifying it once more as
a point load (source for velocity), propagating it backwards in time from $z_r$ to
$z_s$, and finally reading off the velocity trace, interpreted as a
pressure evolution source on $z_s$.

@@

Figure \ref{fig:preddinvhshh0} shows the pressure source gather
produced from the pressure data gather in Figure \ref{dsrcphh0} by
application of the operator on the right-hand side of formula
\ref{eqn:lamident} to the source surface pressure gather depicted in
\ref{fig:dsrcphh0}. The homogeneous model is used in all propagations
implicit in the prescription \ref{eqn:lamident}. The scale is the same
as for Figures \ref{fig:dinvhshh0} and \ref{fig:dhshh0}. In fact
Figure \ref{fig:preddinvhshh0} appears to be nearly identical to
Figure \ref{fig:dinvhshh0}, and both appear to be dip-filtered
versions of Figure \ref{fig:dhshh0}. The difference plot (Figure
  \ref{fig:ddiffinvhshh0}, using the same color scale, shows that the
  similarity is quantitative. The recovered source gather (Figure
  \ref{fig:preddinvhshh0}) also generates the same acoustic fields as
  the point source at $z=z_d$: the pressure gather at $z=z_r=1000$ m
  depth is shown in Figure \ref{fig:dpredhsrecphh0}, and its
  difference with the point source gather (Figure \ref{fig:drecphh0})
  plotted on the same color scale in Figure
  \ref{fig:ddiffpredhsrecphh0}.

@@

To illustrate these points, I have re-computed the source gather
estimates shown in Figures \ref{fig:preddinvhshh0} and
\ref{fig:preddinvhsll0}. In place of the receiver geometry in the
target data (Figures \ref{fig:drecphh0} and \ref{fig:drecpll0}, I
place the receiver array at $z_s+\Delta z=2900$ m depth, just 100 m above the
source surface at $z_s=3000$ m. For the discretization used to create
the examples shown so far, that is just a 5 gridpoint difference in
depth, as opposed to 100 gridpoints between the source and receiver
depths for examples such as shown in Figure \ref{fig:recphh0} or
\ref{fig:drecphh0}. The resulting approximation to $\Lambda_{z_s}^+$
via equation \ref{eqn:lamident} for this configuration is evaluated in
the same way as before in Figures \ref{fig:preddnshshh0}, \ref{fig:ddiffnshshh0},
\ref{fig:dprednshsrecphh0}, and \ref{fig:ddiffprednshsrecphh0}
(homogeneous model) and Figures \ref{fig:preddnshsll0}, \ref{fig:ddiffnshsll0},
\ref{fig:dprednshsrecpll0}, and \ref{fig:ddiffprednshsrecpll0} (lens
model). The effect of aperture limitation is clearly diminished: the
second figure in each series compares the full-aperture pressure source
gathers (Figures \ref{fig:dhshh0.su} and \ref{fig:dhsll0.su} with the
approximated image of the corresponding pressure gather (Figure
\ref{fig:dsrcphh0}) under the approximation to $\Lambda_{z_s}^+$, and
the last figures show that the approximated source gathers accurately
predict the point-source pressure gather at the receiver datum
$z_r=1000$ m.
As mentioned in the last section, computation of the transpose of $\Lambda^+$ (exact, not approximate in
the high frequency sense) is critical to the successful
construction of the preconditioner. The relation \ref{eqn:lamident}
does not provide a computation for this operator. However set
\begin{equation}
  \label{eqn:lamtilde}
  \tilde{\Lambda}^+_{z_s} = -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
    S}^+_{z_s,z_s+\Delta z}\Pi_1^T.
\end{equation}
Then \ref{eqn:lamnear} can be rewritten
\[
  \Lambda^+_{z_s} \approx \tilde{\Lambda}^+_{z_s}.
\]
Of course, all of the examples so far show images of
$\tilde{\Lambda}^+_{z_s}$.

@@

Figure \ref{fig:symmdnshshh0} shows the output of the symmetrized
approximate source-to-pressure operator per equation \ref{eqn:wcomp},
applied once again to the pressure data in Figure
\ref{fig:dsrcp}. Note the resemblance to Figures
\ref{fig:preddnshshh0} and \ref{fig:preddnshstrhh0}. These are all
asymptotic approximations of each other.

@@

@@

@@

@@

>>>>

Next observe that in view of the relation \ref{eqn:tracejump20}, and
the downgoing nature of the ray system carrying the high frequency
energy in $(p^+,\bv^+)$, the field $(\tilde{p}^-,\tilde{\bv}^-)$ is
actually the restriction to $z<z_r$ of the anti-causal solution of \ref{eqn:awepm}
with $z_s$ replaced by $z_r$, zero constitutive defect, and vertical
load given by the jump in pressure at $z=z_r$ - for this field, use
the same notation. Continuity of vertical
velocity $\tilde{v}^-_z$ at $z=z_r$ implies that the vertical load is
\[
  f_r = -[\tilde{p}^-]|_{z=z_r} =-(\lim_{z\rightarrow
    z_r^+}\tilde{p}^- - \lim_{z\rightarrow
    z_r^-}\tilde{p}^-)
\]
\[
  \approx 2 P_r \tilde{p}^- = 2 P_r p^+
\]
(from the definition \ref{eqn:defsamp}, $P_r$ is the limit from the
left). Thus
\[
  P_s \tilde{v^-_z} \approx V^-_{z_r,z_s}(2 P_rp^+) \approx
  2V^-_{z_r,z_s}S^+_{z_r,z_s}h_s.
\]
so
\[
  h_s \approx -2 P_s v^+_z \approx -2 P_s \tilde{v}^-_z \approx
  -4V^-_{z_r,z_s}S^+_{z_r,z_s}h_s
\]
Combine this observation with \ref{eqn:tildevtohsubs} to obtain
\[
 -4  V^-_{z_r,z_s} S^+_{z_s,z_r}  \approx  I,
\]
This relation combines with the identity \ref{eqn:trtrcomp} to
yield the first main result of this section:
\begin{eqnarray}
  \label{eqn:approxinv}
  (V^+_{z_s,z_r})^T S^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
  (S^+_{z_s,z_r})^T V^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
  V^+_{z_s,z_r} (S^+_{z_s,z_r})^T & \approx & \frac{1}{4}I, \nonumber\\
  S^+_{z_s,z_r} (V^+_{z_s,z_r})^T & \approx & \frac{1}{4}I.
\end{eqnarray}.
The second equation is simply the transpose of the first, and the
last two follow by by an exactly analogous argument using time
reversal and interchange of the roles of $z_s$ and$z_r$.

@@

Figure \ref{fig:dinvhshh0} shows the approximate inversion (via the
first equation in display \ref{eqn:approxinv}) of the
pressure gather shown in Figure \ref{fig:drecphh0}, inverted in the homogeneous model
used to generate the data. The result is a scaled, dip-filtered
version of the pressure source shown in Figure \ref{fig:dhshh0}.
The dip filter effect is unavoidable and is caused by the aperture
limitation of the acquisition geometry: the steeper dips in the source
gather (Figure \ref{fig:dhshh0}) do not contribute to the data, and are not
recovered in the inversion.

@@

The second feature is a decoupling of the models used (implicitly or
otherwise) in simulation and those used in inversion: it is possible
to {\em accurately invert pressure data for pressure source using the
  wrong model}, in the sense that the inverted source will
generate an accurate recovery of the input data re-simulated in the
same (wrong) model. For example, inversion of the data of the previous
example (Figure \ref{fig:drecplh0}) {\em in the homogeneous model}
results in an inverted source gather at $z=z_s$ (Figure \ref{fig:dinvhslh0}
that is very far from a dip-filtered version of the source used
generate the data (Figure \ref{fig:dhshh0}). However, this result is an
accurate inversion: re-simulation {\em using the same (homogeneous)
  model as used in the inversion} results in accurate recovery of the
input pressure gather (Figure \ref{fig:drecplh0}). The difference is
shown on the same color scale in Figure \ref{fig:ddiffrecplh0}.

@@

Relation \ref{eqn:tracejump10} characterizes
$\Lambda^+_{z_s}$ as connecting the pressure and velocity components
of downgoing fields restricted to $z=z_s$. That is, the pressure
source gather
$h_s=\Lambda^{+}_{z_s}P_s p^+ = -2
P_{z_s}v_z$, displayed in Figure \ref{fig:dhshh0}, is the image of the
pressure gather in Figure \ref{fig:dsrcphh0} under
$\Lambda^{+}_{z_s}$.
The pressure gather
\ref{fig:dfwdphh0} is the image of this pressure source gather under
$S^+_{z_s,z_r}$. The corresponding vertical velocity gather
(Figure \ref{fig:dfwdvzhh0}) is $-1/2$ times $\Lambda^+_{z_r}P_r
p$. Therefore scaling the data in Figure \ref{fig:dfwdvzhh0} by
$-2$ produces
$\Lambda^+_{z_r}S^+_{z_s,z_r}\Lambda^{+}_{z_s}P_s p^+$. On the other
hand, figure \ref{fig:daltvzhh0} shows the result of applying
$V^+_{z_s,z_r}$ to $f_s=-2P_s p^+$. Therefore scaling the gather in
Figure \ref{fig:daltvzhh0} by $-\frac{1}{2}$ produces $V^+_{z_s,z_r}P_zp^+$.
Since the data in Figures \ref{fig:dfwdvzhh0} and \ref{fig:daltvzhh0}
are essentially identical, the relation \ref{eqn:sv} holds for this
example.

As shown in the last section, $4(V_{z_s,z_r}^+)^T$ is approximately
inverse to $S^{+}_{z_s,z_r}$. Therefore, transposing both sides of
equation \ref{eqn:sv} and using \ref{eqn:approxinv}, obtain
\begin{equation}
  \label{eqn:almostunitary}
  4(V_{z_s,z_r}^+)^TS^+_{z_s,z_r} = [ (\Lambda^+_{z_s})^T
  (S^{+}_{z_s,z_r})^T(\Lambda^+_{z_r})^T]S^{+}_{z_s,z_r} \approx I.
\end{equation}

The remarkable feature of the identity \ref{eqn:almostunitary} is that
it exhibits an approximate right inverse of $S^+$ as an adjoint with
respect to a weighted inner product - or it would, if the operators
$(\Lambda^+)$ were symmetric positive definite. As noted earlier,
these operators are only approximately symmetric, though they are
positive semi-definite. That is not a great obstacle, however:
symmetrizing them in the obvious way commits a negligible error, of
the sort that this paper already neglects wholesale. That is,
\begin{equation}
  \label{eqn:unitary}
  [ \frac{1}{2}((\Lambda^+_{z_s})^T+ \Lambda^+_{z_s})
  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_r})^T+
  \Lambda^+_{z_r})]S^{+}_{z_s,z_r} \approx I.
\end{equation}

To illustrate this unitary property of $S^+_{z_s,z_r}$, I apply the
operator
\[
  \frac{1}{2}((\Lambda^+_{z_s})^T+ \Lambda^+_{z_s})
  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_r})^T+
  \Lambda^+_{z_r})
\]
to the data $S^+_{z_s,z_r}h_s$ (Figure \ref{fig:drerecphh0}), in which
$h_s$ is the dip-filtered version (Figure \ref{fig:dinvhshh0}) of the
downgoing source created earlier (Figure \ref{fig:dhshh0}), produced
by inversion of the data in Figure \ref{fig:drecphh0} via the first
equation in display \ref{eqn:approxinv}, propagation in the homogenous
model.  The operator above is computed via the technique explained in
the section ``Computing and Symmetrizing $\Lambda$'', below, using
auxiliary receiver arrays 100 m above the data source and receiver arrays.

The output is shown in Figure
\ref{fig:lamsstlamrdrecphh0}. The difference with the actual source is
shown in Figure \ref{fig:difflamsstlamrdrecphh0}. A similar exercise
using the lens data \ref{fig:drecplh0} but inversion in the homogenous
model produces the extended source depicted in
\ref{fig:lamsstlamrdrecplh0}, which differs from the inversion result
via equation \ref{eqn:approxinv} by the gather displayed in Figure
\ref{fig:difflamsstlamrdrecplh0}.

The symmetrized $\Lambda$ operators are at least positive
semi-definite, hence define (at least) semi-norms.
Similar relations have been derived for other scattering operators,
and have been used to accelerate iterative solutions of inverse
scatering problems: \cite{DafniSymes:SEG18b} review some of this
literature.

@@

For convenience, in this section write $S$ in place of
$S^+_{z_s,z_r}$. Also abbreviate the symmetrized $\Lambda$ operators
using notation suggesting weight
operators in model and data spaces:
\begin{eqnarray}
  W_m^{-1}&=& \frac{1}{2}((\Lambda^+_{z_s})^T+
              \Lambda^+_{z_s}),\nonumber \\
  W_d &=& \frac{1}{2}((\Lambda^+_{z_r})^T+ \Lambda^+_{z_r}).
          \label{eqn:wdef}
\end{eqnarray}
The identification of the symmetrized $\Lambda^+_{z_s}$ as the inverse
of another operator $W_m$ is formal, since the former operator is
likely to have null (or nearly-null) vectors due to aperture-related
amplitude loss. Since some version of $W_m$ is essential in the
formulation for effective preconditioning, I will derive a usable
candidate to stand in for it below.

@@

Note that the normal operator appearing on the left-hand side of
\ref{eqn:norm0} is not an approximate identity, due to the presence of
the regularization term: the spectrum increases in spread with
increasing $\alpha$, leading to slower convergence. Fortunately for
the present setting, the operators $W_m^{-1}$, $A$, and $W_m$
approximately commute (they are scalar {\em pseudodifferential}, once
the difficulties with the definition of $W_m$, mentioned above, are
taken care of), so that $A^{\dagger} \approx A^T$. Therefore
\begin{equation}
  \label{eqn:normapprox}
  S^{\dagger}S + \alpha^2 A^{\dagger}A \approx I + \alpha^2A^TA
\end{equation}
Recall that $A$ is simply multiplication by the Euclidean distance to
the physical source point $\bx_s$: $A u (\bx) = |\bx-\bx_s|u(\bx),
A^TAu(\bx) = |\bx-\bx_s|^2u(\bx)$. So the equation $(I+\alpha^2
A^TA)u=b$ is trivial to solve, and this is a key characteristic of a
good preconditioner. However this observation must be combined with
the weighted norm structure.

Rewrite the normal equation \ref{eqn:norm0} as
\begin{equation}
  \label{eqn:norm1}
  W_m^{-1}(S^TW_dS + \alpha^2 A^TW_mA)h = W_m^{-1}S^TW_md 
\end{equation}
Since $W_m$ is self-adjoint and positive definite, the common factor on both sides of \ref{eqn:norm1} can be re-written as
\begin{equation}
  \label{eqn:normpart}
  Nh = (S^*S + \alpha^2 A^*A)h = S^*d 
\end{equation}
in which $S^*, A^*$ are the adjoints with the original (Euclidean)
inner product in the domains but the weighted inner product in data
space:
\begin{eqnarray}
  \label{eqn:sadj}
  S^* &=& S^T W_d,\\
  A^* &=& A^T W_m.
\end{eqnarray}
Note the $S^*S$ and $A^*A$ are symmetric in the Euclidean sense, so
equation \ref{eqn:normpart} is a symmetric positive (semi-)definite
linear system, just the sort of thing for which the 
The Preconditioned Conjugate Gradient (``PCG'') algorithm was
designed. PCG for solution
of equation \ref{eqn:normpart} with preconditioner $M$is usually
written as Algorithm 1 (see for example \cite{Golub:2012}):

@@

Computations of $\Lambda^{\pm}_{z_s}$ and its transpose are clearly critical steps in an
implementation of the PCG algorithm outlined in the preceding section.
Direct computation of the pressure-to-source operator $\Lambda^{\pm}_{z_s}$, for instance by solving
\ref{eqn:awepm} and reading off $P_sv^{\pm}_{z}$, turns out to be
numerically ill-behaved. The relation \ref{eqn:snull} provides and
alternative approach, taking advantage of the accurate approximate
inverse to $S^+_{z_s,z_r}$ constructed above. The first row of
\ref{eqn:snull}, slightly rearranged, is
\begin{equation}
  \label{eqn:lamidea}
  \Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T f_s \approx  -\frac{1}{2} S^+_{z_s,z_r}\Lambda^+_{z_s}f_s.
\end{equation}
The approximate inverse construction for $S^++_{z_s,z_r}$ permits (approximate)
solution of this equation for $\Lambda^+_{z_s}f_s$: apply
$4(V^{+}_{z_s,z_r})^T$ to both sides of equation \ref{eqn:lamidea} and
use the first equation in the list \ref{eqn:approxinv} to get
\begin{equation}
  \label{eqn:lamident}
  \Lambda^+_{z_s} \approx -8(V^{+}_{z_s,z_r})^T\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T.
\end{equation}
This identity is the major result of this section: it shows how
to compute that action of $\Lambda^+_{z_s}$ by propagating the input
pressure trace, identified as a source for the velocity evolution,
forward in time from $z_s$ to $z_r$
reading off the pressure trace on $z=z_r$, identifying it once more as
a point load (source for velocity), propagating it backwards in time from $z_r$ to
$z_s$, and finally reading off the velocity trace, interpreted as a
pressure evolution source on $z_s$.

@@

Figure \ref{fig:preddinvhshh0} shows the pressure source gather
produced from the pressure data gather in Figure \ref{fig:dsrcphh0} by
application of the operator on the right-hand side of formula
\ref{eqn:lamident} to the source surface pressure gather depicted in
\ref{fig:dsrcphh0}. The homogeneous model is used in all propagations
implicit in the prescription \ref{eqn:lamident}. The scale is the same
as for Figures \ref{fig:dinvhshh0} and \ref{fig:dhshh0}. In fact
Figure \ref{fig:preddinvhshh0} appears to be nearly identical to
Figure \ref{fig:dinvhshh0}, and both appear to be dip-filtered
versions of Figure \ref{fig:dhshh0}. The difference plot (Figure
  \ref{fig:ddiffinvhshh0}), using the same color scale, shows that the
  similarity is quantitative. The recovered source gather (Figure
  \ref{fig:preddinvhshh0}) also generates the same acoustic fields as
  the point source at $z=z_d$: the pressure gather at $z=z_r=1000$ m
  depth is shown in Figure \ref{fig:dpredhsrecphh0}, and its
  difference with the point source gather (Figure \ref{fig:drecphh0})
  plotted on the same color scale in Figure
  \ref{fig:ddiffpredhsrecphh0}.

@@

To illustrate these points, I have re-computed the source gather
estimates shown in Figure \ref{fig:preddinvhshh0}. In place of the receiver geometry in the
target data (Figure \ref{fig:drecphh0}), I
place a receiver array at $z_s+\Delta z=2900$ m depth, just 100 m above the
source surface at $z_s=3000$ m. For the discretization used to create
the examples shown so far, that is just a 5 gridpoint difference in
depth, as opposed to 100 gridpoints between the source and receiver
depths for examples such as shown in Figure \ref{fig:recphh0} or
\ref{fig:drecphh0}.

Note that the homogeneous and lens models are identical in the depth
range $2900 < z < 3000$ m, so the computed $\Lambda^+_{z_s}$ operators
will be the same for both models. Hence I show only results for the
homogenous model.

The approximation to $\Lambda_{z_s}^+$ via equation \ref{eqn:lamident}
for this configuration is evaluated in the same way as before in
Figures \ref{fig:preddnshshh0}, \ref{fig:ddiffnshshh0},
\ref{fig:dprednshsrecphh0}, and \ref{fig:ddiffprednshsrecphh0}. The effect of aperture
limitation is clearly diminished: the second figure in this series
compares the full-aperture pressure source gather (Figure
\ref{fig:dhshh0}) with the 
image of the corresponding pressure gather (Figure \ref{fig:dsrcphh0})
under the approximation to $\Lambda_{z_s}^+$, and the last figures
show that the approximated source gathers accurately predict the
point-source pressure gather at the receiver datum $z_r=1000$ m.

As mentioned in the last section, computation of the transpose of
$\Lambda^+$ (exact, not approximate in the high frequency sense) is
critical to the successful construction of the preconditioner. The
relation \ref{eqn:lamident} does not provide a computation for this
operator. However set
\begin{equation}
  \label{eqn:lamtilde}
  \tilde{\Lambda}^+_{z_s} = -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
    S}^+_{z_s,z_s+\Delta z}\Pi_1^T.
\end{equation}
Then \ref{eqn:lamnear} can be rewritten
\[
  \Lambda^+_{z_s} \approx \tilde{\Lambda}^+_{z_s}.
\]
Of course, all of the examples so far show images of
$\tilde{\Lambda}^+_{z_s}$.

@@

Figure \ref{fig:symmdnshshh0} shows the output of the symmetrized
approximate source-to-pressure operator per equation \ref{eqn:wcomp},
applied once again to the pressure data in Figure
\ref{fig:dsrcphh0}. Note the resemblance to Figures
\ref{fig:preddnshshh0} and \ref{fig:preddnshstrhh0}. These are all
asymptotic approximations of each other.

@@

\plot{lamsstlamrdrecphh0}{width=\textwidth}{Inversion of data shown in
  Figure \ref{fig:drecphh0}, simulated in homogeneous model, using the
  approximate unitarity relation \ref{eqn:unitary} and propagation in
  homogenous model.}

\plot{difflamsstlamrdrecphh0}{width=\textwidth}{Difference between
  data displayed in Figures \ref{fig:dinvhshh0} and
  \ref{fig:lamsstlamrdrecphh0}, plotted on the same color scale.}

\plot{lamsstlamrdrecplh0}{width=\textwidth}{Inversion of data shown in
  Figure \ref{fig:drecplh0}, simulated in lens model, using the
  approximate unitarity relation \ref{eqn:unitary} and propagation in
  homogenous model.}

\plot{difflamsstlamrdrecplh0}{width=\textwidth}{Difference between
  data displayed in Figures \ref{fig:dinvhslh0} and
  \ref{fig:lamsstlamrdrecplh0}, plotted on the same color scale.}

@@

\plot{dpredhsrecphh0}{width=\textwidth}{Pressure gather at receiver 
  datum $z=z_r=1000$ m simulated in homogeneous model from source 
  gather shown in Figure \ref{fig:preddinvhshh0}, also based on the 
  homogeneous model. Compare with point source pressure gather (Figure 
  \ref{fig:drecphh0}).}

\clearpage

\plot{ddiffpredhsrecphh0}{width=\textwidth}{Plot of difference 
  between data shown in Figures \ref{fig:drecphh0} and 
  \ref{fig:dpredhsrecphh0}, plotted on same color scale as the latter 
  two figures.}

@@

\plot{dpredhsrecpll0}{width=\textwidth}{Pressure gather at receiver 
  datum $z=z_r=1000$ m simulated in lens model from source 
  gather shown in Figure \ref{fig:preddinvhsll0}, also based on the 
  homogeneous model. Compare with point source pressure gather (Figure 
  \ref{fig:drecplh0}).}

\plot{ddiffpredhsrecpll0}{width=\textwidth}{Plot of difference 
  between data shown in Figures \ref{fig:drecplh0} and 
  \ref{fig:dpredhsrecpll0}, plotted on same color scale as the latter 
  two figures.}

\clearpage

@@

\plot{compreshh0}{width=\textwidth}{Comparison of residual Euclidean 
  norms: CG (blue), PCG (red), plotted vs. iteration. Data =
  homogeneous model, point source (Figure \ref{fig:recphh0}), inversion 
  in homogeneous model.}

\plot{compreslh0}{width=\textwidth}{Comparison of residual Euclidean 
  norms: CG (blue), PCG (red), plotted vs. iteration. Data =
  lens model, point source (Figure \ref{fig:recplh0}), inversion 
  in homogenous model.}

\plot{compresll0}{width=\textwidth}{Comparison of residual Euclidean 
  norms: CG (blue), PCG (red), plotted vs. iteration. Data =
  lens model, point source (Figure \ref{fig:recplh0}), inversion 
  in lens model.}
